Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-05-22 appendix.tex
Adding a clarifying sentence, and clearing out :+1: comments.
update form argument and add proptos

Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, if there is a unique maximum of the log-likelihood we will have $x_1=x_2$ at the maximum.
An analogous statement holds for $y_1$ and $y_2$.
It remains to show that the joint inference procedure maximizing \eqref{eq:profile_likelihood} results in a unique estimate of $t$.
Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
We decompose the likelihood into the divergence and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.
Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$; we restrict our attention to $x_1,y_1,x_2,y_2\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
%EM This looks much better now. Can we get a subscript on tau for these calculations?
%das: I changed all of them that were already subscripts, but did you also want subscripts on ones that didn't have them, e.g., D'' and g_1, etc.?
Define
\[
g_j(x_1) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \{x_1,y_1,x_2,y_2,w\}),
\]
i.e., the generating probability of the $j$th site split as a function of $x_1$.
Recall the divergence
\[
\shannonDivergence_{\tau_1,t^*}(\tau_1,\tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}),
\]
which, simplifying notation for clarity, we define as a function of $x_1$ as
\[
D(x_1) = \sum_{j=1}^\nSiteSplits c_j \cdot \log g_j(x_1)
\]
where $c_j \ge 0$ for $j=1,\ldots,\nSiteSplits$.
It remains to show $D''(x_1) < 0$ for $x_1,y_1,x_2,y_2\in(0,1)$.
By straightforward calculus,
\[
D''(x_1) = \sum_{j=1}^\nSiteSplits c_j \cdot \left(\frac{g_j(x_1)g_j''(x_1) - g_j'(x_1)^2}{g_j(x_1)^2}\right).
\]
Using each functional form in Table~\ref{tab:gen-sitepatprob} we see $g_j''(x_1)=0$ and $g_j(x_1)>0$ for $x_1,y_1,x_2,y_2\in(0,1)$ and $j=1,\ldots,\nSiteSplits$ so that
\[
D''(x_1) = -\sum_{j=1}^\nSiteSplits c_j \cdot \left(\frac{g_j'(x_1)}{g_j(x_1)}\right)^2.
\]
Finally, we require at least one $j$ where $c_j \neq 0$ and $g_j'(x_1) \neq 0$ so that $D''(x_1) \not\equiv 0$.
Consider the case $j=1$ where $\siteSplit_1=\emptyset$ so that, ignoring the $1/8$ constant in each term,
\[
g_1(x_1) \propto 1 + x_1x_2 +  y_1y_2 +  w[  x_1 + x_2][  y_1 + y_2] + x_1y_1x_2y_2,
\]
\[
g_1'(x_1) \propto x_2 + y_1w + y_2w + y_1x_2y_2,
\]
and
\[
c_1 \propto 1+(x^*)^2+(y^*)^2+4x^*(y^*)^2+(x^*)^2(y^*)^2.
\]
We have $c_1 > 0$ by $x^*,y^*\in[0,1]$ and, by assumption, $g_1'(x_1) > 0$ for $x_1,y_1,x_2,y_2\in(0,1)$.
This implies the above second derivative is negative.

We now focus on the partial log-likelihood term.
Recall the partial log-likelihood takes the form
\[
\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau_1, \tilde{t}).
\]
As above, define the partial log-likelihood as a function of $x_1$ as
\[
\tilde{\ell}(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_j(x_1)
\]
where
\[
h_j(x_1) = \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau_1, \{x_1,y_1,x_2,y_2,w\}).
\]
Doing calculation as in Figure~\ref{fig:example_likelihoods} with $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, we see each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, will be
\[
h_j(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
\]
for $e_j\in\{0,1\}$, which, simplifying, results in
\begin{align}
\tilde{\ell}(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1).
\end{align}
This has a second derivative
\[
\tilde{\ell}''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
\]
Again, by assumption $x_1,y_1,x_2,y_2\in(0,1)$, and we need only $0 \le \sum_j c_je_j \le 1$ so that, by the second derivative test, the sum of $D(x_1)$ and $\tilde{\ell}(x_1)$ has a unique maximum.
%EM Shouldn't you at least mention that the second deriv isn't exactly zero as you did before?
%das: added a sentence
Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\tilde{\ell}''(x_1) \le 0$.
Thus, with $D''(x_1) < 0$, the sum $D''(x_1) + \tilde{\ell}''(x_1) < 0$, implying $D(x_1) + \tilde{\ell}(x_1)$ is concave as a function of $x_1\in(0,1)$, and there exists a single $x_1\in(0,1)$ maximizing the likelihood.
Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

p_{12} = p_{23} = 1-x^2-y^2+x^2y^2 & = (1+x)(1-x)(1+y)(1-y) \\
p_{1} = p_{3} = 1-x^2+y^2-x^2y^2 & = (1+x)(1-x)(1+y^2) \le (1+x)(1-x)(1+y) \\
p_{2} = p_{123} = 1+x^2-y^2-x^2y^2 & = (1+x^2)(1+y)(1-y) \le (1+x)(1+y)(1-y).

@@

    p_{\emptyset} & = \log(1+x^2+y^2+4xyw+x^2y^2) \\
                  & \le \log(1+x^2+y^2+4xy+x^2y^2) \\
                  & \le \log(2(1+x)(1+y))

@@

    p_{13} & = \log(1+x^2+y^2-4xyw+x^2y^2) \\
                  & \le \log(1+x^2+y^2+x^2y^2) \\
                  & \le \log((1+x)(1+y)).

@@

Theorem~\ref{thm:general-bl} allows us to see where inconsistency in branch length estimation arises when performing general joint maximization over all of the branch parameters, though its conclusion is a statement about both the inconsistency of the internal branch estimate as well as about poor estimation of the other branch parameter estimates.
To obtain an analytic result indicating where only the internal branch parameter is inconsistently estimated, we consider a restricted case of Theorem~\ref{thm:general-bl}.
Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and estimate the internal branch parameter $w$.
%das: @EM I made this not a theorem since it's now a special case. Should it stay a theorem?
Let
\[
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
\]
\[
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
\]
The maximum likelihood value $\hat{w}$ is equal to 1 if
\begin{equation}
\label{eq:restricted-bl-result}
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
-\gamma^2\left(1 + \frac{1}{2}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0,
\end{equation}
and there exists a set of $0 < x^*, y^* < 1$ satisfying this.
This follows in a straightforward manner from \eqref{eq:general-bl-result}.
If $\hat{x}=x^*$ and $\hat{y}=y^*$, then $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$ and \eqref{eq:general-bl-result} becomes \eqref{eq:restricted-bl-result}.
This allows us to demarcate a region of biased internal branch length estimation by plotting where the inequality in \eqref{eq:restricted-bl-result} is satisfied (Fig.~\ref{fig:bl-inconsistency}).

>>>>

Since exchanging $x_1$ and $x_2$ does not change the value of the log-likelihood $\ell_{\tau_1,t^*}(\tau_1, \tilde{t})$, if there is a unique maximum of the log-likelihood we will have $x_1=x_2$ at the maximum.
An analogous statement holds for $y_1$ and $y_2$.
It remains to show that the joint inference procedure maximizing \eqref{eq:profile_likelihood} results in a unique estimate of $t$.
We show the Hessian of the joint likelihood term is negative definite, which implies there is a unique maximizer of \eqref{eq:profile_likelihood}.
We decompose the likelihood into the divergence and partial log-likelihood terms as in \eqref{eq:log_likelihood_simplified}.

%Without loss of generality, we focus on the single parameter $x_1$ and show we obtain a unique maximum when performing joint inference; similar arguments hold for the remaining parameters.
Consider the general case of $\tilde{t}=\{x_1, y_2, x_1, y_2, w\}$; we restrict our attention to $x_1,y_1,x_2,y_2\in(0,1)$, i.e., cases where $0<p_{\siteSplit_j}<1$ for all $\siteSplit_j$.
Recall the divergence
\[
\shannonDivergence_{\tau_1,t^*}(\tau_1,\tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}).
\]
By Gibbs inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, \tilde{t}) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$.
Referring to Tables~\ref{tab:gen-sitepatprob} and~\ref{tab:sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
-(x^*)^2 = -x_1x_2+y_1w(x_1-x_2)
\]
and for $\{1\}$ and $\{1,2\}$ yields
\[
-(x^*)^2 = -x_1x_2+y_2w(x_2-x_1).
\]
The difference of these two equations simplifies to
\[
0 = (y_1w+y_2w)(x_1-x_2)
\]
and, since $y_1w+y_2w > 0$ by assumption, we have $x_1=x_2$.
This along with the previous equalities imply $x_1=x_2=x^*$.
We use the same process to show that, at the maximum, $y_1=y_2=y^*$, and thus that the divergence term attains a unique maximum over $x_1,y_1,x_2,y_2\in(0,1)$.

We now focus on the partial log-likelihood term.
Recall the partial log-likelihood takes the form
\[
\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t^*)\cdot\log \Pr(\ancestralSplitRV=\xi_j \mid \siteSplitRV = \siteSplit_j, \tau_1, \tilde{t}).
\]
The partial log-likelihood can be decomposed into a sum of functions of each variable, i.e.,
\[
\tilde{\ell}_{\tau_1,t^*}(\tau_1, \tilde{t}) = \sum_{j=1}^\nSiteSplits c_j\cdot\left(\log h_{j,1}(x_1)+\log h_{j,2}(y_1)+\log h_{j,3}(x_2)+\log h_{j,4}(y_2)+\log h_{j,5}(w)\right).
\]
Due to this additive form, any off-diagonal terms of the Hessian for this function will be zero, so it remains to show that the diagonal terms are nonpositive.
Without loss of generality we focus on the variable $x_1$ and the partial log-likelihood proportional to
\[
\tilde{\ell}(x_1) = \sum_{j=1}^\nSiteSplits c_j\cdot\log h_{j,1}(x_1).
\]
Doing calculation as in Figure~\ref{fig:example_likelihoods} with $\tilde{t}=\{x_1,y_1,x_2,y_2,w\}$, we see each functional form, suppressing constants with respect to $x_1$ and the initial $1/32$ constant, will be
\[
h_{j,1}(x_1) \propto (1+x_1)^{e_j}(1-x_1)^{1-e_j}
\]
for $e_j\in\{0,1\}$, which, simplifying, results in
\begin{align}
\tilde{\ell}(x_1) &\propto \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(\sum_{j=1}^\nSiteSplits c_j(1-e_j)\right) \log (1-x_1)\\
                  &= \left(\sum_{j=1}^\nSiteSplits c_je_j\right) \log(1+x_1) + \left(1-\sum_{j=1}^\nSiteSplits c_je_j\right) \log (1-x_1).
\end{align}
This has a second derivative
\[
\tilde{\ell}''(x_1) = -\left(\frac{\sum_j c_je_j}{(1+x_1)^2}+\frac{1-\sum_j c_je_j}{(1-x_1)^2}\right).
\]
Again, by assumption $x_1,y_1,x_2,y_2\in(0,1)$, and we need only $0 \le \sum_j c_je_j \le 1$ so that the diagonal term of the Hessian will be nonpositive.
Since $\sum_j c_j = 1$ and $e_j\in\{0,1\}$, then $0 \le \sum_j c_je_j \le 1$ and $\tilde{\ell}''(x_1) \le 0$.
Applying similar arguments to the other variables, the Hessian for the partial log-likelihood will have nonpositive diagonal terms and off-diagonal terms equal to zero.
Thus, the Hessian of the sum of the divergence and the partial log-likelihood will be negative definite, as the Hessian for the divergence is negative definite, and there exists a unique maximum to the likelihood over $x_1,x_2,y_1,y_2\in(0,1)$.
Given this and the symmetry of the likelihood, we let $x_1=x_2=x$ and $y_1=y_2=y$.
The Felsenstein tree does not admit this property, but, since we are interested in a lower bound for this tree, we simplify the objective function by constraining $x_1=x_2$ and $y_1=y_2$ similarly.

@@

p_{12} = p_{23} \propto 1-x^2-y^2+x^2y^2 & = (1+x)(1-x)(1+y)(1-y) \\
p_{1} = p_{3} \propto 1-x^2+y^2-x^2y^2 & = (1+x)(1-x)(1+y^2) \le (1+x)(1-x)(1+y) \\
p_{2} = p_{123} \propto 1+x^2-y^2-x^2y^2 & = (1+x^2)(1+y)(1-y) \le (1+x)(1+y)(1-y).

@@

    p_{\emptyset} & \propto 1+x^2+y^2+4xyw+x^2y^2 \\
                  & \le 1+x^2+y^2+4xy+x^2y^2 \\
                  & \le 2(1+x)(1+y)

@@

    p_{13} & \propto 1+x^2+y^2-4xyw+x^2y^2 \\
                  & \le 1+x^2+y^2+x^2y^2 \\
                  & \le (1+x)(1+y).

@@

Theorem~\ref{thm:general-bl} allows us to see where inconsistency in branch length estimation arises when performing general joint maximization over all of the branch parameters, though its conclusion is a statement about both the inconsistency of the internal branch estimate as well as about poor estimation of the other branch parameter estimates.
To obtain an analytic result indicating where only the internal branch parameter is inconsistently estimated, we consider a restricted case of Theorem~\ref{thm:general-bl}.
Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and estimate the internal branch parameter $w$.
Let
\[
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
\]
\[
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
\]
The maximum likelihood value $\hat{w}$ is equal to 1 if
\begin{equation}
\label{eq:restricted-bl-result}
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
-\gamma^2\left(1 + \frac{1}{2}\left(1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2\right)\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0,
\end{equation}
and there exists a set of $0 < x^*, y^* < 1$ satisfying this.
This follows in a straightforward manner from \eqref{eq:general-bl-result}.
If $\hat{x}=x^*$ and $\hat{y}=y^*$, then $\gamma_{L} = \gamma_{U} = \gamma$ and $\beta_{L} = \beta$ and \eqref{eq:general-bl-result} becomes \eqref{eq:restricted-bl-result}.
This allows us to demarcate a region of biased internal branch length estimation by plotting where the inequality in \eqref{eq:restricted-bl-result} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
